---
title: "ENVS-193DS_homework-03"
author: "Aiden Soller"
format: html
editor: visual
toc: true # use this to display a table of contents
execute:
  message: false # use this to make sure messages don't show up
  warning: false # use this to make sure warnings don't show up
---

[GitHub repository](https://github.com/aidensoller-glitch/ENVS-193DS_winter-2026_final.git)

```{r}
#| message: false
#| warning: false
library(tidyverse)

library(janitor)

library(gt)

library(car)

library(rstatix)

library(dplyr)

library(here)

library(MuMIn)

library(DHARMa)

library(ggeffects)

kelp <- read_csv("../Data/Annual_Kelp_All_Years_20250903.csv")

mopl <- read_csv("../DATA/MOPL_nest-site_selection_Duchardt_et_al._2019.csv")
```

# Problem 1. Research writing

# a. Transparent statistical methods

In part 1, they used linear regression. We know this because the analysis examined whether a continuous variable (precipitation) can be used to predict another continuous variable (the extent of flooded wetland area).

In part 2, they used a Kruskal–Wallis test. This test evaluates whether the median flooded wetland area differs among several groups, in this case the five water year classifications (wet, above normal, below normal, dry, and critical drought) to determine if at least one group’s median is significantly different from the others.

# b. Figure needed

A boxplot would be an appropriate figure to accompany the test in part 2 because it visually compares distributions and medians across multiple groups, which matches the goal of the Kruskal–Wallis test. On the x-axis I would have the 5 different water year classifications, while on the y-axis I would show the wetland flooding area. Each category on the x-axis would have its own boxplot representing the distribution of flooded wetland area values for that water year type. In each boxplot, there would be a horizontal line inside the box that represents the median flooded wetland area. The box shows the interquartile range, while whiskers would extend to the rest of the non-outlier data. This figure would help readers visually compare the medians and variability of flooded wetland area across the five water year classifications, providing clearer context for the statistical test.

# c. Suggestions for rewriting

Part 1: Flooded wetland area in the San Joaquin River Delta did not show evidence of being predicted by precipitation. This suggests that differences in yearly precipitation were not linked to measurable changes in the extent of flooded wetlands in this dataset. (Linear regression, F(df) = \[test statistic\], p = 0.11, α = significance level).

Part 2: Flooded wetland area did not appear to differ across water year classifications, indicating that the median flooded wetland area was similar among the five water year categories in this analysis. (Kruskal–Wallis test: H = \[test statistic\], p = 0.12, α = significance level).

# d. Interpretation

One additional variable that could influence flooded wetland area is river discharge (water flow) entering the delta. River discharge is a continuous variable, and higher flow rates could increase the amount of water entering wetlands, potentially expanding the total flooded area.

# Problem 2. Data visualization

# a. Cleaning and summarizing

```{r}
kelp_clean <- kelp |> # store the kelp data in kelp_clean 
  clean_names() |> # clean column names
  
  filter(site %in% c("IVEE", "NAPL", "CARP")) |> # Include observations from Isla Vista, Naples, and Carpinteria
  
  filter(!(date == "2000-09-22" & transect == "6" & quad == "40")) |> # removes observation
  
  mutate(site_name = case_when( # Create descriptive site_name based on site codes
  site == "NAPL" ~ "Naples", # change site name from NAPL to Naples
  site == "IVEE" ~ "Isla Vista", # change site name from IVEE to Isla Vista
  site == "CARP" ~ "Carpinteria" # change site name from CARP to Carpinteria
)) |> 
  
mutate(site_name = as_factor(site_name), # converts site_name to a factor
       site_name = fct_relevel(site_name, # custom order for factor levels
                               "Naples", "Isla Vista", "Carpinteria")) |>
  
  group_by(site_name, year) |> # Group by site_name and year for summarizing
  
  summarize(mean_fronds = mean(fronds)) |> # Calculate mean fronds per group
  ungroup() # Ungroup to remove grouping, giving data frame

kelp_clean |> slice_sample(n = 5) # slice 5 samples from the kelp_clean data frame 


```

```{r}
str(kelp_clean) # Show the structure of dataset
```

# b. Understanding the data

That observation was excluded because the fronds value for 2000-09-22 in transect 6, quad 40 is -99999, which is not an actual count but a message indicating missing or unrecorded data. The data catalog explains this in the Methods and Protocols tab on the data page, under the method's section.

# c. Visualize the data

```{r}
ggplot(kelp_clean, aes(x = year, 
                       y = mean_fronds, 
                       color = site_name)) +
  
  geom_line(size = 1) + # line geometry
  geom_point(size = 2) + # point geometry
  
  scale_x_continuous(
    breaks = c(2000, 2005, 2010, 2015, 2020, 2025),
    minor_breaks = seq(2000, 2025, 2.5)
  ) +
  
  scale_y_continuous(
    breaks = c(0, 10, 20, 30, 40),
    minor_breaks = seq(0, 40, 5)
  ) +
  
  scale_color_manual(values = c(
    "Naples" = "darkcyan",
    "Isla Vista" = "goldenrod",
    "Carpinteria" = "thistle"
  )) +
  
  labs(
    title = "Sites vary in mean kelp fronds per year",
    x = "Year",
    y = "Mean kelp fronds per year",
    color = "Site"
  ) +
  
  theme_minimal(base_family = "Monaco") +
  
  theme(
    plot.title = element_text(hjust = 0.5),
    panel.border = element_blank(),
    panel.grid.major = element_line(
      color = "grey75",
      linewidth = 0.6),
    panel.grid.minor = element_line(
      color = "grey80",
      linewidth = 0.4)
  )
```

# d. Interpretation

The site with the most variable mean kelp fronds per year is Isla Vista, as seen through its large fluctuations ranging from 4 to 40 mean kelp fronds per year. Additionally, Isla Vista's fluctuation's are more variable year over year than both Naples and Carpinteria with large increases and decreases in the mean kelp fronds as seen in years like 2006-2008 or 2022-2023.

The site with the least variable mean kelp fronds per year is Carpinteria, as seen through its comparitvely small fluctuations ranging only from 3-16 mean kelp fronds per year. While similiar, Naples ranges to 16 mean kelp fronds per year 3 times, whereas Carpinteria only ranges that high once.

# Problem 3. Data analysis

# a. Response variable

In the data set 0s mean that a site within a prairie dog colony is unused and available, whereas a 1 means that a site is used as a Mountain plover nest site.

# b. Purpose of study

Visual obstruction is a continuous variable, as it represents a numeric value representing vegetation height and density using a Robel pole. As visual obstruction increases, the probability of Mountain Plover nest site use would likely decrease because the species prefers areas with short vegetation and more open ground that allow better visibility of predators such as the environment seen at prairie dog colonies. The discription of visual obstruction and the use of a Robel pole is found in the methods section, "Adult Mountain Plover Density – Data Collection,” third paragraph, whereas the justification of the decrease in probability of a nest is found in the discussion section, paragraph 4.

Distance to prairie dog colony edge is a continuous variable because it represents a numeric distance from a nest site to a prairie dog colony. As the distance to prairie dog colony edge increase, the probability of Mountain Plover nest site use would decrease because prairie dog colonies provide the environment that Mountain Plovers prefer by shortening the vegation on areas with open ground. The information was found in the discussion section, paragraph 3.

# c. Table of models

| Model number | Visual Obstruction | Distance to Prarie Dog Colony | Model Description |
|----------------|:--------------:|:-------------------:|:------------------:|
| 0 | N | N | no predictors (null model) |
| 1 | Y | Y | all predictors (saturated model) |
| 2 | N | Y | Distance to colony edge only |
| 3 | Y | N | Visual obstruction only |

# d. Run the models

```{r}
mopl_clean <- mopl |>
  clean_names() |> # clean column names
  mutate(used_bin = case_when(id == "used" ~ 1, # mutate collumn "used" = 1, "available" = 0
                              id == "available" ~ 0)) |>
  select(used_bin, edgedist, vo_nest) # keep only needed columns

m0 <- glm( # fit a generalized linear model
  used_bin ~ 1, # null model (no predictors)
  data = mopl_clean, # use mopl_clean data set
  family = "binomial" # logistic regression for nest used (1) or not used (0)
)

m1 <- glm( # fit a generalized linear model
  used_bin ~ vo_nest + edgedist, # saturated model (all predictors)
  data = mopl_clean, # use mopl_clean data set
  family = "binomial" # logistic regression for nest used (1) or not used (0)
)

m2 <- glm( # fit a generalized linear model
  used_bin ~ edgedist, # Distance to colony edge predictor
  data = mopl_clean, # use mopl_clean data set
  family = "binomial" # logistic regression for nest used (1) or not used (0)
)

m3 <- glm( # fit a generalized linear model
  used_bin ~ vo_nest, # Visual obstruction predictor
  data = mopl_clean, # use mopl_clean data set
  family = "binomial" # logistic regression for nest used (1) or not used (0)
)

```

# e. Select the best model

```{r}
AICc(m0, m1, m2, m3) |> # calculate AICc values for all four models
  arrange(AICc) # sort models from lowest (best) to highest AICc
```

The best model as determined by Akaike's Information Criterion (AIC) is the model which used visual obstruction as the only predictor variable to the probability of site usage by Mountain Plover nests.

# f. Check the diagnostics

```{r}
sim_res_m3 <- simulateResiduals(fittedModel = m3) # # simulate residuals for the best model m3 and store as sim_res_m3 
plot(sim_res_m3) # create plot of residuals
```

# g. Visualize the model predictions

```{r}
predict_data_m3 <- ggpredict(
  m3,
  terms = "vo_nest"
)

ggplot(data=mopl_clean, # use mopl_clean dataset
       aes(x = vo_nest, # set vo_nest as x-axis
           y = used_bin)) + # set used_bin as y-axis
  geom_point( # plot observed points
    color = "blue", # edit line color
    alpha = 0.5 ) + # edit point transparency
  
  geom_ribbon(data = predict_data_m3, # add confidence interval
              aes(x = x,
                  y = predicted,
                  ymin = conf.low,
                  ymax = conf.high),
              fill = "skyblue", # edit fill color
  ) +
  geom_line(data = predict_data_m3, # add prediction line
           aes(x = x,
               y = predicted),
           color = "red" # edit line color
  ) +
  labs(
    x = "Visual Obstruction (cm)", # x-axis label
    y = "Probability of Mountain Plover nest site use (Yes = 1, No = 0)" # y-axis label
  ) + 
  theme_classic() # change theme to classic
  
```

# h. Write a caption for your figure

# Figure 1. Predicted probability of Mountain Plover nest site use in relation to visual obstruction.

Blue points show observed nest site use (0 = unused, 1 = used) across levels of visual obstruction (cm). The red line represents predicted probabilities from the selected logistic regression model, and the light blue shaded ribbon indicates the 95% confidence interval around those predictions. Data collected by Duchardt, Courtney; Beck, Jeffrey; Augustine, David (2020). Data from: Mountain Plover habitat selection and nest survival in relation to weather variability and spatial attributes of Black-tailed Prairie Dog disturbance \[Dataset\]. Dryad.

<https://doi.org/10.5061/dryad.ttdz08kt7>

# i. Calculate model predictions
```{r}
ggpredict(m3, # use the selected logistic regression model
          terms = "vo_nest [0]") # set visual obstruction value to 0
```
# j. Interpret your results

Figure 1 shows that the probability of Mountain Plover nest site use declines as visual obstruction increases, indicating that plovers prefer areas with more open ground. There is a clear negative relationship between vegetation height and the likelihood that a site will be used for nesting. When visual obstruction is zero, the predicted probability of a site being used for nesting is about 0.73. This pattern likely reflects the plovers’ need for unobstructed sightlines to detect predators and monitor their surroundings. Additionally, open areas may provide easier access to food and safer conditions for raising chicks, making sites with taller or denser vegetation less suitable for nesting.
